### 2025-04-15 causal-inference.qmd
20250409 - minor tweaks
20250415 - reorganization

### Causation

@@

The first criterion for establishing causality involves temporal precedence.
In order for a cause to influence an effect, the cause must occur before the effect.
For instance, if sports drink consumption influences player performance, the sports drink consumption (that is presumed to influence performance) must occur prior to the performance improvement.
Establishing the first criterion eliminates the possibility that the association between the purported cause and effect reflects reverse causation.
Reverse causation occurs when the purported effect is actually the cause of the purported cause, rather than the other way around.
For instance, if sports drink consumption occurs only once, and it occurs only before and not after performance, then we have ruled out the possibility of reverse causation (i.e., that better performance causes players to consume sports drink).

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., overestimation) in associations between variables due to [confounding](#sec-causalDiagramConfounding) (i.e., common causes) [@Digitale2022].
For instance, from a DAG, it is possible to determine what variables it is important to control for (and how)—and just as importantly, what not to control for—in order to get unbiased estimates of the effect of one variable on another variable or of the association between two variables of interest.
To create DAGs, you can use the `R` package `dagitty` [@R-daggity; @Textor2016_packages] or the associated browser-based extension, DAGitty: <https://dagitty.net> (archived at <https://perma.cc/U9BY-VZE2>).
Examples of various causal diagrams that could explain why `X` is associated with `Y` are in Figures [-@fig-XCausesY], [-@fig-YCausesX] and [-@fig-ZCausesXandY].

>>>>

## Causation {#sec-causation}

@@

The first criterion for establishing causality involves temporal precedence.
In order for a cause to influence an effect, the cause must occur before the effect.
For instance, if sports drink consumption influences player performance, the sports drink consumption (that is presumed to influence performance) must occur prior to the performance improvement.
Establishing the first criterion eliminates the possibility that the association between the purported cause and effect reflects reverse causation.
*Reverse causation* occurs when the purported effect is actually the cause of the purported cause, rather than the other way around.
For instance, if sports drink consumption occurs only once, and it occurs only before and not after performance, then we have ruled out the possibility of reverse causation (i.e., that better performance causes players to consume sports drink).

@@

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., overestimation) in associations between variables due to [confounding](#sec-causalDiagramConfounding) (i.e., common causes) [@Digitale2022].
For instance, from a DAG, it is possible to determine what variables it is important to control for (and how)—and just as importantly, what not to control for—in order to get unbiased estimates of the effect of one variable on another variable or of the association between two variables of interest.
To create DAGs, you can use the `R` package `dagitty` [@R-dagitty; @Textor2016_packages] or the associated browser-based extension, DAGitty: <https://dagitty.net> (archived at <https://perma.cc/U9BY-VZE2>).
Examples of various causal diagrams that could explain why `X` is associated with `Y` are in Figures [-@fig-XCausesY], [-@fig-YCausesX] and [-@fig-ZCausesXandY].

### 2025-04-15 correlation.qmd
20250415 - reorganization

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("corrplot")
library("correlation")
library("tidyverse")
```

@@

@@

### Covariance {#sec-correlationExamplesCovariance}

@@

### Pearson Correlation {#sec-correlationExamplesPearson}

@@

### Scatterplot with Best-Fit Line

@@

### Rank Correlation: Spearman's Rho ($\rho$) {#sec-correlationExamplesSpearman}

@@

### Partial Correlation {#sec-correlationExamplesPartial}

@@

### Nonlinear Correlation {#sec-correlationExamplesNonlinear}

@@

### Correlation Matrix {#sec-correlationExamplesMatrix}

@@

### Correlogram {#sec-correlationExamplesCorrelogram}

>>>>

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("corrplot")
library("correlation")
library("ggridges")
library("tidyverse")
```

@@

### Player Height and Fantasy Points {#sec-correlationHeightFantasyPoints}

@@

#### Covariance {#sec-correlationExamplesCovariance}

@@

#### Pearson Correlation {#sec-correlationExamplesPearson}

@@

#### Scatterplot with Best-Fit Line

@@

#### Rank Correlation: Spearman's Rho ($\rho$) {#sec-correlationExamplesSpearman}

@@

#### Partial Correlation {#sec-correlationExamplesPartial}

@@

#### Nonlinear Correlation {#sec-correlationExamplesNonlinear}

@@

#### Correlation Matrix {#sec-correlationExamplesMatrix}

@@

#### Correlogram {#sec-correlationExamplesCorrelogram}

### 2025-04-15 multiple-regression.qmd
20250415 - reorganization

Those are some of the key assumptions of multiple regression.
However, there are additional assumptions of multiple regression, including ones discussed in the chapter on [Causal Inference](#sec-causalInference).
For instance, the variables included should reflect causal process such that the [predictor variables](#sec-correlationalStudy) influence the [outcome variable](#sec-correlationalStudy), and not the other way around.
That is, the [outcome variable](#sec-correlationalStudy) should not influence the [predictor variables](#sec-correlationalStudy) (i.e., there should be no reverse causation).
In addition, it is important control for any [confound(s)](#sec-causalDiagramConfounding).
If a [confound](#sec-causalDiagramConfounding) is not controlled for, this is called omitted-variable bias, and it leads the researcher to incorrectly attribute the effects of the omitted variable to the included variables.

@@

@@

Let's first examine the bivariate association of each using a scatterplot to evaluate for any potential nonlinearity, as depicted in @fig-scatterplots.

@@

First, let's estimate a multiple regression model with only linear terms:

@@

>>>>

Those are some of the key assumptions of multiple regression.
However, there are additional assumptions of multiple regression, including ones discussed in the chapter on [Causal Inference](#sec-causalInference).
For instance, the variables included should reflect causal process such that the [predictor variables](#sec-correlationalStudy) influence the [outcome variable](#sec-correlationalStudy), and not the other way around.
That is, the [outcome variable](#sec-correlationalStudy) should not influence the [predictor variables](#sec-correlationalStudy) (i.e., there should be no reverse causation).
In addition, it is important control for any [confound(s)](#sec-causalDiagramConfounding).
If a [confound](#sec-causalDiagramConfounding) is not controlled for, this is called *omitted-variable bias*, and it leads the researcher to incorrectly attribute the effects of the omitted variable to the included variables.

@@

#### Examine Descriptive Statistics {#sec-multipleRegressionFantasyPointsWRsDescriptive}

Let's first examine [descriptive statistics](#sec-descriptiveStatistics) of the predictor and outcome variables.

```{r}
player_stats_seasonal %>% 
  dplyr::select(fantasyPoints, age, height, weight, target_share) %>% 
  dplyr::summarise(across(
      everything(),
      .fns = list(
        n = ~ length(na.omit(.)),
        missingness = ~ mean(is.na(.)) * 100,
        M = ~ mean(., na.rm = TRUE),
        SD = ~ sd(., na.rm = TRUE),
        min = ~ min(., na.rm = TRUE),
        max = ~ max(., na.rm = TRUE),
        range = ~ max(., na.rm = TRUE) - min(., na.rm = TRUE),
        IQR = ~ IQR(., na.rm = TRUE),
        MAD = ~ mad(., na.rm = TRUE),
        median = ~ median(., na.rm = TRUE),
        pseudomedian = ~ DescTools::HodgesLehmann(., na.rm = TRUE),
        mode = ~ petersenlab::Mode(., multipleModes = "mean"),
        skewness = ~ psych::skew(., na.rm = TRUE),
        kurtosis = ~ psych::kurtosi(., na.rm = TRUE)),
      .names = "{.col}.{.fn}")) %>%
    tidyr::pivot_longer(
      cols = everything(),
      names_to = c("variable","index"),
      names_sep = "\\.") %>% 
    tidyr::pivot_wider(
      names_from = index,
      values_from = value)
```

Let's also examine the distributions of the variables using a [density plot](#sec-ridgelinePlot), as depicted in @fig-densityPlotMultipleRegressionVars.

```{r}
#| layout: [[100], [50, 50], [50, 50]]
#| label: fig-densityPlotMultipleRegressionVars
#| fig-cap: "Density Plot of Model Variables (i.e., Predictor and Outcome Variables)."
#| fig-alt: "Density Plot of Model Variables (i.e., Predictor and Outcome Variables)."
#| fig-subcap:
#|   - "Fantasy Points"
#|   - "Age"
#|   - "Height"
#|   - "Weight"
#|   - "Target Share"

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("WR")),
  mapping = aes(
    x = fantasyPoints)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    color = "#000000",
    fill = "#0099F8"
  ) +
  geom_density(
    color = "#000000",
    fill = "#F85700",
    alpha = 0.6 # add transparency
  ) +
  geom_rug() +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Fantasy Points with Histogram and Rug Plot"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("WR")),
  mapping = aes(
    x = age)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    color = "#000000",
    fill = "#0099F8"
  ) +
  geom_density(
    color = "#000000",
    fill = "#F85700",
    alpha = 0.6 # add transparency
  ) +
  geom_rug() +
  labs(
    x = "Age (years)",
    y = "Density",
    title = "Density Plot of Player Age with Histogram and Rug Plot"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("WR")),
  mapping = aes(
    x = height)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    color = "#000000",
    fill = "#0099F8"
  ) +
  geom_density(
    color = "#000000",
    fill = "#F85700",
    alpha = 0.6 # add transparency
  ) +
  geom_rug() +
  labs(
    x = "Fantasy Points",
    y = "Height (inches)",
    title = "Density Plot of Player Height with Histogram and Rug Plot"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("WR")),
  mapping = aes(
    x = weight)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    color = "#000000",
    fill = "#0099F8"
  ) +
  geom_density(
    color = "#000000",
    fill = "#F85700",
    alpha = 0.6 # add transparency
  ) +
  geom_rug() +
  labs(
    x = "Weight (pounds)",
    y = "Density",
    title = "Density Plot of Player Weight with Histogram and Rug Plot"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position_group %in% c("WR")),
  mapping = aes(
    x = target_share)
) +
  geom_histogram(
    aes(y = after_stat(density)),
    color = "#000000",
    fill = "#0099F8"
  ) +
  geom_density(
    color = "#000000",
    fill = "#F85700",
    alpha = 0.6 # add transparency
  ) +
  geom_rug() +
  labs(
    x = "Target Share",
    y = "Density",
    title = "Density Plot of Target Share with Histogram and Rug Plot"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

@@

Then, let's examine the bivariate association of each using a scatterplot to evaluate for any potential nonlinearity, as depicted in @fig-scatterplots.

@@

Now that we have examined [descriptive statistics](#sec-descriptiveStatistics) and bivariate associations, let's first estimate a multiple regression model with only linear terms:

@@

We could also generate the model-implied prediction of fantasy points for any value of the [predictor variables](#sec-correlationalStudy).
For instance, here are the number of fantasy points expected for a Wide Receiver who is 23 years old, who is 6'2" tall (72 inches), and weighs
